---
title: "Hitmontop Project"
subtitle: "Proposal"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
library(skimr)
```

# Data 1

## Problem or question

-   Identify the problem you will solve or the question you will answer
-   Explain why you think this topic is important.
-   Identify the types of data/variables you will use.
-   State the major deliverable(s) you will create to solve this problem/answer this question.

**Problem or Question:**

The problem we aim to address in this project is to analyze the factors affecting air quality in a specific urban area using data analysis techniques. We want to understand how various factors, such as weather conditions and traffic, impact air quality. This topic is important because it can provide insights into the drivers of air pollution and assist in developing effective mitigation strategies to improve air quality.

**Types of Data/Variables:**

1.  **Air Quality Data:** We will use air quality measurements, such as PM2.5 and PM10 concentrations, CO2 levels, and ozone levels as our primary variables. These will serve as our primary indicators of air quality.

2.  **Weather Data:** Weather variables like temperature, humidity, wind speed, and precipitation can influence air quality. We will analyze historical weather data to examine these relationships.

3.  **Traffic Data:** Data on traffic patterns, such as traffic volume and congestion, can be a significant factor in air quality. We will incorporate data related to traffic in the area.

4.  **Geospatial Data:** Geographic information system (GIS) data to determine the proximity to major roads and industrial areas.

**Major Deliverables:**

1.  **Exploratory Data Analysis (EDA):** We will conduct EDA to gain insights into the relationships between air quality, weather, and traffic data.

2.  **Data Visualization:** Creating visualizations to communicate findings and patterns effectively to a broader audience.

3.  **Report:** We will produce a comprehensive report summarizing our findings, and conclusions based on the analysis.

## Introduction and data

If you are using a dataset:

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

**Introduction and Data:**

For this project, we will utilize publicly available air quality data from the UCI Machine Learning Repository. The data was originally collected by the Department of Information Engineering and Computer Science at the University of Trento, Italy. The dataset consists of air quality data from an urban area in Italy.

**Source of Data:**

-   **Data Source:** UCI Machine Learning Repository

-   **Original Data Curator:** University of Trento, Italy

-   **Data Collection Period:** The data collection period spans several years and includes air quality measurements.

**Description of Observations:**

The dataset contains time-series data with records of various air quality measurements, weather conditions, and traffic-related data. Each record provides information on specific dates and times, air quality indices, meteorological parameters, and traffic data. Here is a brief overview of some key columns in the dataset:

-   Date and Time: Timestamps for each record.

-   Air Quality Data: Measurements of various air pollutants, including PM2.5, PM10, CO2, and ozone levels.

-   Weather Data: Meteorological variables such as temperature, humidity, wind speed, and precipitation.

-   Traffic Data: Information on traffic, including traffic volume and traffic congestion.

## Glimpse of data

```{r}
#| label: load-data-1

# add code here

data1 <- read.csv("data/AirQualityUCI.csv", sep=";")
glimpse(data1)
```

# Data 2

## Problem or question

-   Identify the problem you will solve or the question you will answer
-   Explain why you think this topic is important.
-   Identify the types of data/variables you will use.
-   State the major deliverable(s) you will create to solve this problem/answer this question.

**Problem or question:**

The problem we aim to address in this data science project is to analyze and understand factors influencing student performance in higher education institutions. We want to investigate the relationship between various student attributes and their academic success, as measured by their final grades or GPAs. This is an important topic because it can help educational institutions and policymakers make informed decisions to support students and improve the overall quality of education.

**Types of data/variables:**

We will use a dataset containing information about individual students' demographics, socio-economic backgrounds, and academic performance. The variables we plan to use include:

1.  Student demographic data (e.g., age, gender, nationality).

2.  Socio-economic factors (e.g., parental education level, family size, economic status).

3.  Academic history (e.g., prior educational achievements, number of past failures).

4.  Study habits and resources (e.g., study time, internet usage).

5.  Final academic outcomes (e.g., final grade or GPA).

    **Major deliverables:**

    To address this problem, we plan to create the following deliverables:

    1.  Exploratory Data Analysis (EDA): We will perform initial data exploration to gain insights into the dataset, visualize key trends, and identify potential relationships between variables.

    2.  Data Visualization: We will create visualizations to present findings effectively, including scatter plots, histograms, and bar charts.

    3.  Report: We will produce a comprehensive report summarizing our findings, and conclusions based on the analysis.

## Introduction and data

If you are using a dataset:

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

**Introduction and data:**

**Source of the data:** The dataset we will use is publicly available on the UCI Machine Learning Repository.

**Original collection:** The data was originally collected by P. Cortez and A. Silva at the University of Minho (Minho, Portugal). It was collected for a research project aimed at understanding the factors that influence student performance in Portuguese schools.

**Description of observations:** The dataset contains information about 649 students from two Portuguese secondary schools. The data includes 30 variables, which encompass a wide range of student attributes and academic performance metrics. This dataset allows us to investigate the factors that may impact student success in secondary education.

## Glimpse of data

```{r}
#| label: load-data-2

# add code here

data2_1 <- read.csv("data/student-mat.csv", sep = ";")
glimpse(data2_1)
data2_2 <- read.csv("data/student-por.csv", sep = ";")
glimpse(data2_2)
```

# Data 3

## Problem or question

-   Identify the problem you will solve or the question you will answer
-   Explain why you think this topic is important.
-   Identify the types of data/variables you will use.
-   State the major deliverable(s) you will create to solve this problem/answer this question.

**Problem or question**

The objective of this project is to analyze Dota2 game results to identify patterns or factors that significantly contribute to a team's victory or defeat. Specifically, the project seeks to understand the correlation between specific game metrics and winning outcomes without delving into predictive modeling or machine learning.

**Importance of the Topic**

Dota 2, being one of the most popular online multiplayer games, not only provides entertainment but also supports a significant e-sports scene. Analyzing game results can provide insights for players, teams, and stakeholders about effective strategies, game-play balance, and factors influencing winning odds. Furthermore, this analysis can contribute to discussions about game improvements and the e-sports industry's growth.

**Types of Data/Variables Used**

The dataset includes various features, primarily:

1.  Game mode (e.g., All Pick, Random Draft, Single Draft)

2.  Game type (e.g., Ranked, Unranked)

3.  Cluster (indicating the server on which the game was played)

4.  Various in-game features (e.g., heroes used, roles played, duration, etc.)

**Major Deliverable(s)**

1.  Exploratory Data Analysis (EDA): We will perform initial data exploration to gain insights into the dataset, visualize key trends, and identify potential relationships between variables.

2.  Data Visualization: We will create visualizations to present findings effectively, including scatter plots, histograms, and bar charts.

3.  Report: We will produce a comprehensive report summarizing our findings, and conclusions based on the analysis.

## Introduction and data

If you are using a dataset:

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

**Introduction and Data**

**Data Source**

The dataset for this analysis is sourced from the UCI Machine Learning Repository, specifically the "Dota2 Games Results" data set.

**Data Collection**

The data was originally collected by the UCI Machine Learning Repository contributors from public Dota 2 API. It contains results of Dota 2 games, including a range of features that describe the match setup such as game mode, game type, and details of the heroes involved in each match.

**Description of the Observations**

The dataset comprises records of individual Dota 2 games, with each observation representing a unique game. Variables include information on the game setup, heroes used by each team, and the game's outcome.

## Glimpse of data

```{r}
#| label: load-data-3

# add code here
data3_1 <- read.csv("data/dota2Test.csv")
glimpse(data3_1)
data3_2 <- read.csv("data/dota2Train.csv")
glimpse(data3_2)
```
